Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2013-05-22 paper.tex
Minor changes to multipole section, and changed name of hydro section.

\subsection{Hydrodynamics and Gravity}

@@

The white dwarf merger problem falls into a class of problems with ``isolated'' boundary conditions, where we assume that all of the mass in the system is enclosed within the physical domain. Analytically, the Poisson equation is customarily solved by assuming that the potential vanishes as one gets very far away from the region of non-zero mass density. However, on a finite computational domain this assumption does not hold in general: there will be some non-zero value of the gravitational potential on the boundaries of the domain. In principle, the Dirichlet boundary values can be computed by doing a direct sum of the mass distribution for each zone outside the domain:
\begin{equation}
  \left. \Phi(\mathbf{x}) \right|_{\text{boundary}} = \sum_{\text{i, j, k}} \frac{G \rho_{\text{ijk}}}{|\mathbf{x} - \mathbf{x}_{\text{ijk}}|}\, \Delta V_{\text{ijk}}.
\end{equation}
If there are $N$ zones per spatial dimension, then there are $\mathcal{O}(N^2)$ boundary zones, and each boundary zone requires a sum over $N^3$ zones, so the direct computation of the boundary conditions is an $\mathcal{O}(N^5)$ operation. However, in a typical simulation we place the boundaries of the domain far away from the region containing most of the mass, so that they do not contaminate the hydrodynamics in the region of interest at the center of the domain. As a result, a common approach is to make some approximation for the mass distribution on the domain that results in a more tractable summation scheme.

@@

The most natural extension of the monopole prescription is to include higher-order multipole moments. If the entire mass distribution is enclosed, then the potential can be expanded as a series of spherical harmonics:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where the origin of the coordinate system is taken to be at the center of the computational domain. The multipole moments can be calculated by expanding the Green's function for the Poisson equation as a series spherical harmonics, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
Some simplification of Equation \ref{spherical_harmonic_expansion} can be achieved by using the addition theorem for spherical harmonics:
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
Here the $P_{l}^{m}$ are the associated Legendre polynomials and the $P_l$ are the Legendre polynomials. After some algebraic simplification, the potential outside of the mass distribution can be written in the following way:
\begin{equation}
  \Phi(\mathbf{x}) \approx \sum_{l=0}^{l_{\text{max}}} \left[Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} + \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right].
\end{equation}
The modified multipole moments are:
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime) d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
Our strategy for the multipole boundary conditions, then, is to pick some value $l_{\text{max}}$ that is of sufficiently high order to capture the distribution of mass on the grid, evaluate the discretized analog of the modified multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m \leq l$, and then directly compute the value of the potential on all of the boundary zones. This is ultimately an $\mathcal{O}(N^3)$ operation, the same order as the monopole approximation, and the wall time required to calculate the boundary conditions will depend on the chosen value of $l_{\text{max}}$.

>>>>

\subsection{Hydrodynamics}

@@

The white dwarf merger problem falls into a class of problems with ``isolated'' boundary conditions, where we assume that all of the mass in the system is enclosed within the physical domain. Analytically, the Poisson equation is customarily solved by assuming that the potential vanishes as one gets very far away from the region of non-zero mass density. However, on a finite computational domain this assumption does not hold in general: there will be some non-zero value of the gravitational potential on the boundaries of the domain. In principle, the Dirichlet boundary values can be computed by doing a direct sum of the mass distribution for each zone outside the domain:
\begin{equation}
  \left. \Phi(\mathbf{x}) \right|_{\text{boundary}} = G\, \sum_{\text{i, j, k}} \frac{G \rho_{\text{ijk}}}{|\mathbf{x} - \mathbf{x}_{\text{ijk}}|}\, \Delta V_{\text{ijk}}.
\end{equation}
If there are $N$ zones per spatial dimension, then there are $\mathcal{O}(N^2)$ boundary zones, and each boundary zone requires a sum over $N^3$ zones, so the direct computation of the boundary conditions is an $\mathcal{O}(N^5)$ operation. However, in a typical simulation we place the boundaries of the domain far away from the region containing most of the mass, so that they do not contaminate the hydrodynamics in the region of interest at the center of the domain. As a result, a common approach is to make some approximation for the mass distribution on the domain that results in a more tractable summation scheme.

@@

The most natural extension of the monopole prescription is to include higher-order multipole moments. If the entire mass distribution is enclosed, then the potential can be expanded as a series of spherical harmonics:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where the origin of the coordinate system is taken to be at the center of the computational domain. The multipole moments can be calculated by expanding the Green's function for the Poisson equation as a series spherical harmonics, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
Some simplification of Equation \ref{spherical_harmonic_expansion} can be achieved by using the addition theorem for spherical harmonics:
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
Here the $P_{l}^{m}$ are the associated Legendre polynomials and the $P_l$ are the Legendre polynomials. After some algebraic simplification, the potential outside of the mass distribution can be written in the following way:
\begin{equation}
  \Phi(\mathbf{x}) \approx \sum_{l=0}^{l_{\text{max}}} \left[Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} + \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right].
\end{equation}
The modified multipole moments are:
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
Our strategy for the multipole boundary conditions, then, is to pick some value $l_{\text{max}}$ that is of sufficiently high order to capture the distribution of mass on the grid, evaluate the discretized analog of the modified multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m \leq l$, and then directly compute the value of the potential on all of the boundary zones. This is ultimately an $\mathcal{O}(N^3)$ operation, the same order as the monopole approximation, and the wall time required to calculate the boundary conditions will depend on the chosen value of $l_{\text{max}}$.
